---
title: "Data Import"
author: "Carolyn Martinez"
date: "2024-09-17"
output: github_document
---
1. Loading pakcages needed 
```{r setup, echo=FALSE, message = FALSE}

library(tidyverse)
library(janitor)
library(readxl)
library(haven)
```

This document will show hoe to import data from R, not form the files.


*How to tell computer where the file is:*

1. Absolute path - leads the computer to the root directory 
   
   Some issues - not efficient

2. Relative path - tells you how to get there from where you are right not 
 
  *Best practice* Do this by creating an Rproject that contains all the moving parts of the project. 

## Importing the data from the files AND clean the column names 
```{r}
litters_df = read_csv("data_import_examples/FAS_litters.csv")

litters_df = janitor::clean_names(litters_df)
```
Note: janitor:: clean_names means "using the function clean_names" in the janitor package

## Look at the data set 

```{r}
litters_df

head(litters_df)
tail(litters_df, 10)
```

##looking at the ENTIRE data frame - opens a new window 
```{r  eval=FALSE}
view(litters_df)
```

## Learning Assessment: loading FAS_pups.csv and cleaning the names 

```{r}
pups_df=read_csv("data_import_examples/FAS_pups.csv")

pups_df = janitor::clean_names(pups_df)

pups_df

```
## How to make data import seamless 
using the ?read_csv command to see the different optns 

*looking at the options*

```{r}
litters_df=
  read_csv(
    file="data_import_examples/FAS_litters.csv",
    col_names = FALSE,
    skip=1
    )
```
  Notes: 
  - skip = 1 allows you to skip lines depending on how the data is structured. 
  - col_names = false is saying the column names are not actually the names. 
  
  
  *when you are missing data*
```{r}
  litters_df=
  read_csv(
  file="data_import_examples/FAS_litters.csv",
  na=c("NA", "", ".")
)
```

*What if we code 'group' as a factor variable? - changing variable types*

```{r}
litters_df=
  read_csv(
  file="data_import_examples/FAS_litters.csv",
  na=c("NA", "", "."),
  col_types = cols(
    Group=col_factor()
    )
  )
```

## Importing an Excel File 

```{r}
mlb_df=read_excel("data_import_examples/mlb11.xlsx", sheet="mlb11")

mlb_df
```
```{r}
pulse_df=read_sas("data_import_examples/public_pulse_data.sas7bdat")

pulse_df

```

## Never use read.csv()

This is not good because it does not give a lot of info on the data 
- missing the type of variables and how the data is coded

## Never do this - you should not be removing the variables from the data -> EVER
```{r}
litters_df$L
```








