---
title: "tidy_data"
author: "Carolyn Martinez"
date: "2024-09-24"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(readxl)
library (haven)
```

This document will show how to clean data.

The data set is not tidy because the same variable is spread accross MANY columns as opposed to just one. 
```{r}
pulse_df = 
  haven::read_sas("data_import_examples/public_pulse_data.sas7bdat") |>
  janitor::clean_names()

pulse_df
```

#Pivot longer 
This data needs to go from wide to long format.

```{r}
pulse_tidy_df = 
  pivot_longer(
    pulse_df, 
    bdi_score_bl:bdi_score_12m,
    names_to = "visit", 
    values_to = "bdi")

pulse_tidy_df
```

The code below will: import, tidy, and transform the PULSE dataset into a usable format. 
```{r}
pulse_df = 
  haven::read_sas("./data_import_examples/public_pulse_data.sas7bdat") |>
  janitor::clean_names() |>
  pivot_longer( 
    bdi_score_bl:bdi_score_12m,
    names_to = "visit", 
    names_prefix = "bdi_score_",
    values_to = "bdi") |>
  mutate(
    visit = replace(visit, visit == "bl", "00m"),
    visit = factor(visit)) 
```
Do one more example. 

```{r learning assessment}
litters_wide = 
  read_csv(
    "./data_import_examples/FAS_litters.csv",
    na = c("NA", ".", "")) |>
  janitor::clean_names() |>
  select(litter_number, ends_with("weight")) |> 
  pivot_longer(
    gd0_weight:gd18_weight,
    names_to = "gd", 
    values_to = "weight") |> 
  mutate(
    gd = case_match(
      gd,
      "gd0_weight"  ~ 0,
      "gd18_weight" ~ 18
    ))
```

#Pivot Wider 
```{r data frame}
analysis_result = 
  tibble(
    group = c("treatment", "treatment", "placebo", "placebo"),
    time = c("pre", "post", "pre", "post"),
    mean = c(4, 8, 3.5, 4)
    
    
  )
```
This is a made up data set. 
- the analysis_df is tidy long format but if the intention is to have humans be able too read/analyse it, you might want to pivot_wider. 


```{r pivot_wider}
pivot_wider(
  analysis_result, 
  names_from = "time", 
  values_from = "mean"
  )|>
  knitr::kable()
```

#Joiimg FAS Data 

```{r}
litter_df = 
  read_csv(
    "./data_import_examples/FAS_litters.csv",
    na = c("NA", ".", "")) |>
  janitor::clean_names() |>
  separate(group, into = c("dose", "day_of_tx"), sep = 3) |>
  relocate(litter_number) |>
  mutate(
    wt_gain = gd18_weight - gd0_weight,
    dose = str_to_lower(dose))
```

Import 'pups' data set

```{r}
pup_df = 
  read_csv("./data_import_examples/FAS_pups.csv", na = c("NA", "", ".")) |>
  janitor::clean_names() |>
  mutate(
    sex = 
      case_match(
        sex, 
        1 ~ "male", 
        2 ~ "female"),
    sex = as.factor(sex)) 
```








